Conceptual and methodological problems with bias detection and avoidance in natural language processing

### 2021-06-08 thesis/_book/_main.tex
revised point 2

\textbf{state the general topic and give some background}

Natural language processing (NLP) is a subfield of computer science that
processes and analyzes language in text and speech with the use of
modern programming methods. It has practical applications in everyday
life as it concerns tasks such as email filters, smart assistants,
search results, language translations, text analytics and so on. Models
used to accomplish these tasks need a lot of data to learn from. This
data originates from humans activities and historical recordings like
texts, messages, speeches. It turns out that in the learning process
these models can learn implicit biases that reflect harmful
stereotypical thinking still present in modern societies. One can find
methods that aim at identifying hidden biases and/or try to remove them
by modifying the models explicitly. There are many different types of
models in NLP depending on a task that they ought to solve. However all
of them need as an input words represented as numerical values and this
is accomplished with word embedding models. The biases seem to have
their primary source in the way the words are assigned the numerical
values. \newline

\textbf{review of the literature related to the topic}

There is a bunch of literature available on the topic of bias detection
and mitigation in NLP models. Bolukbasi2016Man focuses on gender biases
that may be observable while investigating the representation of job
occupations and gender in terms of their assigned numerical values. The
authors apply cosine similarity measurement to investigate the
phenomenon where jobs stereotypically associated with a given gender are
in fact in the model situated closer to this gender.

@@

manzini2019black proposes novel way of using a cosine similarity method
to get the information on assumed resemblance between words. They
investigate an approach that enables to measure the bias for a class
(like gender, religion, race) and express the final result with only one
averaged number. \newline

\textbf{define the terms and scope of the topic}

It is worth noticing the general distinction of biases mentioned in
Caliskan2017Semantics. They refer to the publication concerning Implicit
Association Test Greenwald et al., 1998) where certain baseline of bias
phenomenon was introduced. Namely it seems that humans naturally exhibit
some biases and that not always bring social concern. One can imagine
the intuitive associations between for example insects and flowers, and
the feelings of pleasantness or unpleasantness. In general people would
rather associate flowers with feeling pleasant than insects and this
preference could be named a bias or prejudice in some direction. However
this type of preference does not cause an uproar and it is rather
morally neutral case. Unfortunately there are other biases and
prejudices that directly influence the quality of other people's lives
and therefore they should be taken care of.

One can find a bunch of various definitions trying to capture what bias
and fairness actually are. With the choice of the definition,
implications into the real-life applications may change as well as it
was pointed out in Mehrabi2019Survey. They mark out that there exist
different types of biases, the list if long but among others there are
historical bias, representation bias, measurement bias. This indicates
how complex the process itself is. Without the proper understanding and
awareness of the problem, people are prone to unconsciously sustain this
phenomenon.

@@

The topic of discrimination is entangled with another concept which is
fairness. It is essential to grasp some concepts of fairness to take
them into consideration while designing implementation of some machine
learning model. In Mehrabi2019Survey one may notice that depending on
the context and application different definitions may be applied.
\newline

\textbf{outline the current situation} The most popular methods focus on
comparing the similarity between words from protected groups and those
that are considered to be stereotypical or harmful in some way. One can
find in this group methods such as euclidean distance, dot product or
cosine similarity. There are also other ways to detect the effects of
biases. For example through the investigation of the model performance
on certain tasks that validate if the model returns some values
independently on gender or race or not. \newline

\textbf{evaluate the current situation (advantages/ disadvantages) and identify the gap}

In the currently used methods (like cosine similarity) the values of
similarity are often aggregated in a way that may lead to false
conclusions. For example due to the averaging of values and the lack of
confidence interval information. \newline

\textbf{identify the importance of the proposed research} One can find a
number of articles on negative real-life implications resulting from the
presence of unaddressed biases in the machine learning models. \newline

\textbf{state the research problem/ questions}

In the paper we indicate how current methods used to detect biases in
natural language models are limited in terms of confidence interval.
\newline

\textbf{state the research aims and/or research objectives}

Our research tries to answer the question of how to enhance the current
way in which the bias detection is performed to make sure that it is
methodologically valid. \newline

\textbf{state the hypotheses}

Our hypothesis is that there can be greater understanding of data and
bias implications when confidence interval and Bayesian method are
applied to the methodology. \newline
<!-- \textbf{outline the order of information in the thesis} -->

\textbf{outline the methodology}

To discuss!

@@

Bolukbasi, Chang, Zou, Saligrama, \& Kalai
(\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define similarity
between words as the inner product of their normalized vectors (that is,
they use cosine similarity). They focus on examining what the geometry
of word embedding is in regard to ``he'' and ``she'' words. In other
words, whether the similarity between those concepts and other words
reflects expected gender stereotypes. They test this hypothesis by
investigating whether there is a connection between word embeddings
representing certain professions and words referring to gender. They
also evaluate whether automatically produced analogies between words
reflect the stereotypes as well.

@@

In the article Manzini, Lim, Tsvetkov, \& Black
(\protect\hyperlink{ref-manzini2019black}{2019}) the authors present
different approach towards finding similarities between classes of
words. The authors claim that texts available online are full of direct
or indirect human stereotypes. As a result, word embeddings are prone to
learn and maybe amplify those biases and propagate them further into AI
models that are used for various applications. Cosine distance is used
in the article as a measure to first argue for the existence of
multi-class bias and then to show how through bias mitigation techniques
the bias may be decreased. The main steps in the procedure are as
follows.

@@

>>>>

Natural language processing (NLP) is a subfield of computer science that
processes and analyzes language in text and speech with the use of
modern programming methods. It has practical applications in everyday
life as it concerns tasks such as email filters, smart assistants,
search results, language translations, text analytics and so on. Models
used to accomplish these tasks need a lot of data to learn from. This
data originates from humans activities and historical recordings such as
texts, messages or speeches. It turns out that in the learning process
these models can learn implicit biases that reflect harmful
stereotypical thinking still present in modern societies. One can find
methods that aim at identifying and measuring hidden biases and/or try
to remove them by modifying the models. There are many different types
of models in NLP depending on a task that they are supposed to solve.
However, all of them need as an input words represented by means of
numbers and this is accomplished with word embedding models. The models
usually assign the values based on the context in which the words
appear. It means that the input data can have enormous influence on the
outcome. The biases seem to have their primary source in the way the
words are assigned the numerical values.

There is considerable amount of literature available on the topic of
bias detection and mitigation in NLP models. Bolukbasi, Chang, Zou,
Saligrama, \& Kalai (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
focuses on gender biases that may be observable while investigating the
representation of job occupations and gender in terms of their assigned
numerical values. The authors apply cosine similarity measurement to
investigate the phenomenon where (the vectors corresponding to) words
related to jobs that are stereotypically associated with a given gender
are in fact in the model situated closer to this gender. They also use
analogy tasks to evaluate if the bias is present in the word embedding
model. They check analogies by comparing pairs of word vectors, for
example they search for the word complementing the puzzle: man is to
doctor as woman is to \ldots{}? First they subtract word ``man'' from
word ``woman'' and then they search for the ranked list of other words
pairs that have similar vectors' difference. They also include in the
formula a threshold to ensure that the resulting pairs could not be
randomly picked.

However, as in Nissim, Noord, \& Goot
(\protect\hyperlink{ref-Nissim2019Fair}{2019}) it is pointed out, there
are some limitations of this approach. According to the authors in
practice most of analogies implementations do not return any input
words. This means that it does not make sense to expect the algorithm to
return the same profession for both woman and man. Therefore this method
seems to be limited in terms of bias detection. The other problems
regard for example the choice of pairs and words that are used to detect
the presence of discrimination as it is often subjective and without
proper justification. Additionally the choice of parameter set in
Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
formula to ensure that word pairs are not picked by random, is also not
justified and changing it drastically influences the results.

@@

Manzini, Lim, Tsvetkov, \& Black
(\protect\hyperlink{ref-manzini2019black}{2019}) proposes a novel way of
using cosine similarity to obtain the information on assumed resemblance
between words. They investigate an approach that enables them to measure
the bias for a class (like gender, religion, race) and express the final
result with a single metric.

It is worth noticing the general distinction of biases mentioned in
Islam et al. (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}). They
refer to the publication concerning Implicit Association Test Greenwald
et al., 1998) where certain baseline of bias phenomenon was introduced.
Namely it seems that humans naturally exhibit some biases and that not
always bring social concern. One can imagine the intuitive associations
between for example insects and flowers, and the feelings of
pleasantness or unpleasantness. In general people would rather associate
flowers with feeling pleasant than insects and this preference could be
named a bias or prejudice in some direction. However this type of
preference does not cause an uproar and it is rather morally neutral
case. Unfortunately there are other biases and prejudices that directly
influence the quality of other people's lives and therefore they should
be taken care of.

One can find a bunch of various definitions trying to capture what bias
and fairness actually are. With the choice of the definition,
implications into the real-life applications may change as well as it
was pointed out in Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan
(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}). They mark out that
there exist different types of biases, the list if long but among others
there are historical bias, representation bias, measurement bias. This
indicates how complex the process itself is. Without the proper
understanding and awareness of the problem, people are prone to
unconsciously sustain this phenomenon.

@@

The topic of discrimination is entangled with another concept which is
fairness. It is essential to grasp some concepts of fairness to take
them into consideration while designing implementation of some machine
learning model. In Mehrabi2019Survey one may notice that depending on
the context and application different definitions may be applied.

The most popular methods focus on comparing the similarity between words
from protected groups and those that are considered to be stereotypical
or harmful in some way. One can find in this group methods such as
euclidean distance, dot product or cosine similarity. There are also
other ways to detect the effects of biases. For example through the
investigation of the model performance on certain tasks that validate if
the model returns some values independently on gender or race or not.

In the currently used methods (like cosine similarity) the values of
similarity are often aggregated in a way that may lead to false
conclusions. For example due to the averaging of values and the lack of
confidence interval information.

One can find a number of articles on negative real-life implications
resulting from the presence of unaddressed biases in the machine
learning models.

In the paper we indicate how current methods used to detect biases in
natural language models are limited in terms of confidence interval.

Our research tries to answer the question of how to enhance the current
way in which the bias detection is performed to make sure that it is
methodologically valid.

Our hypothesis is that there can be greater understanding of data and
bias implications when confidence interval and Bayesian method are
applied to the methodology.

@@

Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define
similarity between words as the inner product of their normalized
vectors (that is, they use cosine similarity). They focus on examining
what the geometry of word embedding is in regard to ``he'' and ``she''
words. In other words, whether the similarity between those concepts and
other words reflects expected gender stereotypes. They test this
hypothesis by investigating whether there is a connection between word
embeddings representing certain professions and words referring to
gender. They also evaluate whether automatically produced analogies
between words reflect the stereotypes as well.

@@

In the article Manzini et al.
(\protect\hyperlink{ref-manzini2019black}{2019}) the authors present
different approach towards finding similarities between classes of
words. The authors claim that texts available online are full of direct
or indirect human stereotypes. As a result, word embeddings are prone to
learn and maybe amplify those biases and propagate them further into AI
models that are used for various applications. Cosine distance is used
in the article as a measure to first argue for the existence of
multi-class bias and then to show how through bias mitigation techniques
the bias may be decreased. The main steps in the procedure are as
follows.

@@

\hypertarget{ref-Mehrabi2019Survey}{}
Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
(2019). A survey on bias and fairness in machine learning. \emph{CoRR},
\emph{abs/1908.09635}. Retrieved from
\url{http://arxiv.org/abs/1908.09635}

\hypertarget{ref-Nissim2019Fair}{}
Nissim, M., Noord, R. van, \& Goot, R. van der. (2019). Fair is better
than sensational: Man is to doctor as woman is to doctor. \emph{CoRR},
\emph{abs/1905.09866}. Retrieved from
\url{http://arxiv.org/abs/1905.09866}
